---
title: "Exponential CLT & ToothGrowth Data Analysis - Simulation and Basic Inference"
author: "Riccardo Del Signore"
date: "`r format(Sys.time())`"
output: 
  word_document: 
    toc: true
    toc_depth: 2
---

# Overview

In this report, we answer the following questions:
1. How does the Central Limit Theorem (CLT) apply to the Exponential distribution?
2. Inference on the ToothGrowth dataset.

## Part 1: Exponential CLT

Exp(lamda) = 1/lamda
Var(X) = 1/lamda^2

We are going to set lamda as 0.2 and n as 40
Repeating the experiment 1000 times we are going to explore the effect of the CLT on the Exponential distribution, and how it tends to a Normal distribution.

### Parameters and design simulation
```{r}
set.seed(123) #for reproducibility
 #TODO set simulation pareters
lambda <- 0.2  #rate
n <- 40 #sample size
nsim <- 1000 #number of simulations

#theoretical mean and variance
theor_mean <- 1/lambda
theor_var <- (1/lambda)^2 / n
theor_sd <- sqrt(theor_var)

theor_mean; theor_var; theor_sd
```

### Simulation 

Build vector of means using replicate function.
This function will repeat the expression mean(rexp(n, rate = lambda)) nsim times and store the results in a vector.
```{r}
#TODO Build vector means with lennght nsim
means <- replicate(nsim, mean(rexp(n, rate = lambda)))
head(means)
```
### Theoretical vs empirical mean and variance

```{r}
#TODO empirical mean and variance and sd
sample_mean <- mean(means)
sample_var <- var(means)
sample_sd <- sd(means)

df_confronto <- data.frame(theoretical = c(theor_mean, theor_var, theor_sd),
                          empirical = c(sample_mean, sample_var, sample_sd),
                          row.names = c("mean", "variance", "sd"))
knitr::kable(df_confronto, digits = 3, caption = "Theoretical vs Sample")
```

### Histogram of the means
```{r message=FALSE, warning=FALSE, error=FALSE}
#histogram of the means prob = TRUE
hist(means, prob = TRUE, main = "Sample mean vs Theoretical Mean", xlab = "Means") 
lines(density(means), col = "blue", lwd = 2)
curve(dnorm(x, mean = theor_mean, sd = theor_sd), col = "red", lwd = 2, add = TRUE)
#adding vertical line for the 2 means
abline(v = theor_mean, col = "red", lwd = 1)
abline(v = sample_mean, col = "blue", lwd = 1)
#report nnumbers on the plot
text(x = theor_mean + 0.35, y = 0.15, labels = paste(round(theor_mean, 2)), col = "red")
text(x = sample_mean - 0.30, y = 0.15, labels = paste(round(sample_mean, 2)), col = "blue")
```
### QQ-plot
```{r}
#QQ-plot
qqnorm(means, main = "QQ-plot of the sample means")
```
### Sample variance vs theoretical variance
```{r}
plot(density(means), main = "Sample Variance vs Theoretical Variance", xlab = "Means")
curve(dnorm(x, mean = theor_mean, sd = theor_sd), col = "red", lwd = 2, add = TRUE)
curve(dnorm(x, mean = sample_mean, sd = sqrt(sample_var)), col = "blue", lwd = 2, add = TRUE)
```
