---
title: "Classifying Iight-Lifting Execution Quality with Accelerometer Data"
author: "Kernel236"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: show
---

```{r include=FALSE}
library(caret)
library(ggplot2)
library(dplyr)
library(doParallel)
library(parallel)
library(gridExtra)
```


## Executive summary

I built a multi-class classifier to predict the execution quality of barbell lifts (classe A–E) from Iarable sensors. After cleaning high-NA and near-zero-variance predictors and using stratified k-fold cross-validation, a Random Forest achieved 99.6% CV accuracy. Feature importance suggests belt/forearm accelerations and angular velocities are most predictive.

## Data cleaning

I removed non-predictive metadata (IDs/username/timestamps..), dropped predictors with >95% missingness and near-zero variance. This reduces noise and leakage risk without sacrificing signal.

```{r}
# Read dataset
train_coort <-read.csv(here::here("./data/pml-training.csv"))
test_coort <-read.csv(here::here("./data/pml-testing.csv"))

#rRemouve meta variable not uselful fo rthe models
train_coort <- train_coort |>
  select(-c(X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, new_window, num_window))

test_coort <- test_coort |>
  select(-c(X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, new_window, num_window, problem_id))

#NearzeroVar
nzv <- nearZeroVar(train_coort[, -153]) # saveMetrics = FALSE return just indiex for column with low variance
nzv_test <- nearZeroVar(test_coort)

train_coort <- train_coort[,-c(nzv)]
test_coort <- test_coort[,-c(nzv_test)]
# Remove columns with ≥95% missing values
na_threshold <- 0.95
train_coort <- train_coort[, colMeans(is.na(train_coort)) < na_threshold]
test_coort <- test_coort[, colMeans(is.na(test_coort)) < na_threshold]

X_train <- train_coort[,-ncol(train_coort)]
Y_train <- train_coort[,ncol(train_coort)]

```

## Cross Validation and Model Fitting

All model tuning used k-fold stratified CV within the train set; the holdout was used once for a final unbiased performance estimate.

```{r}
set.seed(123)

ctrl <- trainControl(
  method = "cv",
  number = 5,
  savePredictions = "final",
  classProbs = TRUE
)
```


### Random Forest (RF) and Gradient Boosting (GBM)

I fit a Random Forest as a strong non-parametric baseline. RF handles nonlinearities/interaction and is robust to noisy predictors
```{r fit_models, cache=TRUE, warning=FALSE, message=FALSE}
set.seed(123)

n_cores <- detectCores() # DETECT NUMBER OF CORES
cl <- makeCluster(n_cores -8)
registerDoParallel(cl) # ACTIVE PARALLEL COMPUT ON N - 8 CORE

fit_rf <- train(x = X_train, y = Y_train, method = "rf", trControl = ctrl, tuneLength = 5, metric = "Accuracy")
saveRDS(fit_rf, "../models/fit_rf.rds")

fit_gbm <- train(x = X_train, y = Y_train, method = "gbm", trControl = ctrl, ) #GBM MODEL
saveRDS(fit_rf, "../models/fit_gbm.rds")

stopCluster(cl) #STOP MultiCORE
registerDoSEQ() # RETURN TO 1 by one core
```


### Model rationale

I compared a robust tree-ensemble baseline (Random Forest) with a gradient boosting model (GBM). Ensembles typically perform Ill on noisy, tabular sensor data and are insensitive to feature scaling.

```{r}
res <- resamples(list(rf = fit_rf, gbm = fit_gbm))
dotplot(res)
```
```{r}
fit_rf
```

```{r}
fit_gbm
```

In 5 K-fold Cross validation the RF model show a better Accuracy and Kappa with a very tiny 95% CI

### Features Importance
```{r}
varImp_rf <- varImp(fit_rf)

plot(varImp_rf, top = 20, main = "Top 20 Important Features - Random Forest")
```


## Model Comparison and Expected Error

We trained two models (Random Forest and Gradient Boosting Machine) using 5-fold cross-validation.

- Random Forest achieved an accuracy of 99,6%, corresponding to an expected out-of-sample error of ~0,4%.  
- GBM achieved an accuracy of 96%, corresponding to an expected out-of-sample error of ~4%.  

Based on these results, Random Forest was selected as the final model, since it provides higher accuracy and lower expected error.

## Test cases (20)

I applied the exact same preprocessing to the 20 unlabeled cases and predict using the RF model.

```{r}
pred_rf <- predict(fit_rf, test_coort)
write.csv(pred_rf, file = "../report/pred_rd.csv", row.names = FALSE)
```

